---
title: "Visualization Examples"
author: "Erwan Le Pennec / Eric Matzner"
date: "Spring 2016"
output: 
  html_document: 
    self_contained: no
---

```{r Knitr_Global_Options, include=FALSE}
library("knitr")
opts_chunk$set(warning = FALSE, message = FALSE, cache = TRUE, autodep = TRUE, tidy = FALSE)
library("rgl")
knit_hooks$set(webgl = hook_webgl)
```


# Anscombe quartet

```{r Anscombe}
library("dplyr")
library("ggplot2")

anscombe1 <- transmute(anscombe, example = 1, x = x1, y = y1)
anscombe2 <- transmute(anscombe, example = 2, x = x2, y = y2)
anscombe3 <- transmute(anscombe, example = 3, x = x3, y = y3)
anscombe4 <- transmute(anscombe, example = 4, x = x4, y = y4)

ggplot(data = rbind(anscombe1, anscombe2, anscombe3, anscombe4), aes(x=x,y=y)) +
  geom_point() + geom_smooth(method = "lm", fullrange = TRUE) + 
  facet_wrap( ~ example, ncol = 2)
```

# Playfair

```{r Playfair}
library("HistData")
library("reshape2")
ggplot(data = melt(Wheat, "Year"), aes(x = Year, y = value, color = variable)) + geom_step()

ggplot(data = Wheat, aes(x=Year)) + geom_step(aes(y=Wheat/Wages))

ggplot(data = Wheat, aes(x=Year)) + geom_step(aes(y=Wheat/Wages)) + scale_y_continuous(limits=c(0,NA))
```

# Nightingale
```{r Nightingale}
Nightingale2 <- melt(Nightingale, "Date", c("Wounds", "Other", "Disease"))

ggplot(data = filter(Nightingale2, Date <= "1855-03-02"), aes(x = as.factor(Date))) + geom_bar(aes(y= value, fill = variable), color = "black", width = 1, stat = "identity") + coord_polar(start = -pi/2, direction = 1) + scale_y_sqrt()

ggplot(data = filter(Nightingale2, Date > "1855-03-02"), aes(x = as.factor(Date))) + geom_bar(aes(y= value, fill = variable), color = "black", width = 1, stat = "identity") + coord_polar(start = -pi/2, direction = 1) + scale_y_sqrt()

ggplot(data = filter(Nightingale2, Date <= "1855-03-02"), aes(x = as.factor(Date))) + geom_bar(aes(y= value, fill = variable), color = "black", width = 1, stat = "identity") + coord_polar(start = -pi/2, direction = 1)

ggplot(data = filter(Nightingale2, Date <= "1855-03-02"), aes(x = as.factor(Date))) + geom_bar(aes(y= value, fill = variable), color = "black", width = 1, stat = "identity")

ggplot(Nightingale2, aes(x = as.factor(Date))) + geom_bar(aes(y= value, fill = variable), color = "black", width = 1, stat = "identity")

ggplot(data = summarise(group_by(Nightingale2, variable), value = sum(value)), aes(x = variable, y = value, fill = variable)) + geom_bar(stat = "identity")
```

# Challenger
```{r Challenger}
data(challeng, package = "alr3")

ggplot(data = filter(challeng, Fail > 0), aes(x = Temp, y = Fail)) + geom_point(size = 5)

ggplot(data = filter(challeng, Fail > 0), aes(x = Temp, y = Fail)) + geom_point(size = 5) + geom_smooth(method = "lm", formula = y ~ poly(x,2))

ggplot(data = challeng, aes(x = Temp, y = Fail)) + geom_point(size = 5)

ggplot(data = challeng, aes(x = Temp, y = Fail)) + geom_point(size = 5) + geom_smooth(method = "lm", formula = y ~ poly(x,2))

ggplot(data = challeng, aes(x = Temp, y = Fail)) + geom_point(size = 5) + geom_smooth(method = "lm", formula = y ~ poly(x,2), fullrange = TRUE) + scale_x_continuous(limit = c(32,NA))
```

# Pie plot

```{r Pie}
ggplot(data = diamonds, aes(x = factor(1), fill = cut)) + geom_bar(width = 1) +
  coord_polar(theta = "y") + xlab("") + ylab("") +
  theme(axis.ticks = element_blank(), axis.text.y = element_blank())
```

# Bar

```{r Bar}
ggplot(data = diamonds, aes(x = cut, fill = cut)) + geom_bar()
```

# Cleveland Dot

```{r Cleveland_Dot}
ggplot(data = diamonds, aes(x = cut)) + 
  geom_point(stat = "count", size = 5) +  coord_flip() +
  theme(panel.grid.major.x = element_blank() ,
        panel.grid.major.y = element_line(linetype=3, color="darkgray"))

ggplot(data = diamonds, aes(x = cut, color = cut, fill = cut)) + 
  geom_point(stat = "count", size = 5) +
  geom_bar(width = .01) + coord_flip() +
  theme(panel.grid.major.x = element_blank() ,
        panel.grid.major.y = element_line(linetype=3, color="darkgray"))

ggplot(data = add_rownames(mtcars), aes(x = reorder(rowname,mpg), y = mpg, color = cyl)) + geom_point(size = 3) + coord_flip() + xlab("") +
  theme(panel.grid.major.x = element_blank() ,
        panel.grid.major.y = element_line(linetype=3, color="darkgray"))
```

# Histogram and density

```{r Histo}
ggplot(data = diamonds, aes(x = price)) + geom_histogram()

ggplot(data = diamonds, aes(x = price)) + geom_histogram(binwidth = 15)

ggplot(data = diamonds, aes(x = price)) + geom_density()

ggplot(data = diamonds, aes(x = price)) + geom_density(adjust = .01)
```

# Boxplot and violin plot

```{r Box}
ggplot(data = diamonds, aes(x = factor(1), y = price)) + geom_boxplot()

ggplot(data = diamonds, aes(x = factor(1), y = price)) + geom_violin()
```

# Dot plot

```{r Dotplot}
small_diamonds <- sample_frac(diamonds,.01)

ggplot(data = small_diamonds, aes(x = factor(1), y = price)) +
  geom_dotplot(binaxis = "y", stackdir = "center", stackratio = 1.25, dotsize = .25)

ggplot(data = small_diamonds, aes(x = factor(1), y = price)) +
  geom_dotplot(binaxis = "y", stackdir = "center", stackratio = 1.25, dotsize = .25) +
  stat_summary(fun.y = median, geom = "point", size = 5, color = "red")
```

# Stem

```{r Stem}
stem(small_diamonds[["price"]], scale = 3)
```

# Grouping

```{r Grouping}
ggplot(data = diamonds, aes(x = price, color = cut)) + geom_density()

ggplot(data = diamonds, aes(x = price, fill = cut)) + geom_density(position = "stack")

ggplot(data = diamonds, aes(x = price, color = cut, fill = cut)) +
  geom_density() + facet_wrap(~ cut)

ggplot(data = diamonds, aes(x = cut, y = price)) + geom_violin()

ggplot(data = small_diamonds, aes(x = cut, y = price)) +
  geom_dotplot(binaxis = "y", stackdir = "center", stackratio = 1.25, dotsize = .25) +
  stat_summary(fun.y = median, geom = "point", size = 5, color = "red")

ggplot(data = small_diamonds, aes(x = factor(1), y = price, color = cut)) +
  geom_dotplot(binaxis = "y", stackdir = "center", stackratio = 1.25, dotsize = .25, binpositions = "all", stackgroups =  TRUE) +
  stat_summary(fun.y = median, geom = "point", size = 5, color = "red")
```

# Scatter Plots

```{r Scatter_Plots}
ggplot(data = diamonds, aes(x = carat, y = price)) + geom_point()

ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) + geom_point()

ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) + geom_point(alpha = .25)

ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) + geom_point(alpha = .25) + facet_wrap( ~ cut)
```

# Smoothing

```{r Smoothing}
ggplot(data = diamonds, aes(x = carat, y = price)) + geom_point() +
  geom_smooth()

ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = .25) + geom_smooth()
```

# Decoration

```{r Decoration}
ggplot(data = mtcars, aes(x = hp, y = mpg, size = gear, color = cyl, shape = factor(am))) + geom_point() + scale_size_continuous(range = c(4,8))
```

# Scatter Plot Matrix

```{r Scatter_Plot_Matrix}
library(GGally)
ggpairs(mtcars)
```

# Parallel Coordinates and Radar Plots

```{r Parallel_Coordinates}
# rescale all variables to lie between 0 and 1
scaled <- as.data.frame(lapply(mtcars, ggplot2:::rescale01))
scaled$model <- rownames(mtcars)    # add model names as a variable
mtcarsm <- reshape2::melt(scaled)

ggplot(mtcarsm, aes(x = variable, y = value)) +
  geom_line(aes(group = model, color = model), size = 2) +
  theme(strip.text.x = element_text(size = rel(0.8)),
        axis.text.x = element_text(size = rel(0.8))) +
  guides(color = guide_legend(ncol=2))

ggplot(mtcarsm, aes(x = variable, y = value)) +
  geom_line(aes(group = model, color = model), size = 2) +
  theme(strip.text.x = element_text(size = rel(0.8)),
        axis.text.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  guides(color = "none") + facet_wrap(~ model)
```

```{r Radar_Plots}
coord_radar <- function (theta = "x", start = 0, direction = 1) 
{
    theta <- match.arg(theta, c("x", "y"))
    r <- if (theta == "x") 
        "y"
    else "x"
    ggproto("CordPolar", CoordPolar, theta = theta, r = r, start = start, 
        direction = sign(direction),
        is_linear = function(coord) TRUE)
}

ggplot(mtcarsm, aes(x = variable, y = value)) +
  geom_polygon(aes(group = model, color = model), fill = NA, size = 1, show.legend = FALSE)  +
  coord_radar() +
  facet_wrap( ~ model, nrow = 4) +
  guides(color = guide_legend(ncol=2)) +
  theme(strip.text.x = element_text(size = rel(0.8)),
        axis.text.x = element_text(size = rel(0.8)),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank())

ggplot(mtcarsm, aes(x = variable, y = value)) +
  geom_polygon(aes(group = model, color = model), fill = NA, size = 2, show.legend = FALSE) +
  geom_path(aes(group = model, color = model), size = 2) +
  coord_radar() + 
  theme(strip.text.x = element_text(size = rel(0.8)),
        axis.text.x = element_text(size = rel(0.8))) +
  guides(color = guide_legend(ncol=2))
```

```{r Alluvial}
library(alluvial)
tit <- as.data.frame(Titanic)
alluvial( tit[,1:4], freq=tit$Freq, border=NA,
         hide = tit$Freq < quantile(tit$Freq, .50),
         col=ifelse( tit$Survived == "No", "red", "gray") )
```

# Time Series

```{r Time_Series}
library(quantmod)
tckrs <- c("SPY", "QQQ", "GDX", "DBO", "VWO")
getSymbols(tckrs, from = "2007-01-01")
SPY.Close <- SPY[,4]
QQQ.Close <- QQQ[,4]
GDX.Close <- GDX[,4]
DBO.Close <- DBO[,4]
VWO.Close <- VWO[,4]
SPY1 <- as.numeric(SPY.Close[1])
QQQ1 <- as.numeric(QQQ.Close[1])
GDX1 <- as.numeric(GDX.Close[1])
DBO1 <- as.numeric(DBO.Close[1])
VWO1 <- as.numeric(VWO.Close[1])
SPY <- SPY.Close/SPY1
QQQ <- QQQ.Close/QQQ1
GDX <- GDX.Close/GDX1
DBO <- DBO.Close/DBO1
VWO <- VWO.Close/VWO1
basket <- add_rownames(as.data.frame(cbind(SPY, QQQ, GDX, DBO, VWO)), var = "date")
basket <- mutate(basket, date = as.Date(date))
library("reshape2")
basket_melt <- melt(basket, "date")

ggplot(data = basket_melt, aes(x = date, y = value, color = variable)) + geom_line()

ggplot(data = basket_melt, aes(x = date, y = value, color = variable)) + geom_line() + facet_wrap( ~ variable)
```

# Time Line

```{r Time_Line}
library(timeline)
data(ww2)
timeline(ww2, ww2.events, event.spots=2, event.label="", event.above=FALSE)
```

# Choroplet

```{r Choroplet}
st <- map_data('state')
data(votes.repub)
colnames(votes.repub) <- paste("Y", colnames(votes.repub), sep="")
votes.repub <- mutate(add_rownames(as.data.frame(votes.repub), "region"), region = tolower(region))
st <- left_join(st, votes.repub, by = "region")

ggplot(st, aes(long, lat, group=group, fill = Y1960)) +
    geom_polygon() + scale_fill_continuous(limits = c(20,80))

st_melt <- melt(st, c("long","lat","group","order","region","subregion"))

ggplot(st_melt, aes(long, lat, group=group, fill = value)) +
    geom_polygon() + scale_fill_continuous(limits = c(20,80)) + facet_wrap(~ variable)
```

# Symbol Maps

```{r Symbols_Map}
DecauxKey <- "da879af595184f071c181408b837b7da636f924f"
UrlDecaux <- function(decaux,key) {
  if (grepl('\\?',decaux, perl = TRUE)) {
    delim <- '&'
  }
  else {
      delim <- '?'
  }
  sprintf("https://api.jcdecaux.com/vls/v1/%s%sapiKey=%s",decaux,delim,key)
}

GetJsonDecaux <- function(decaux, key = DecauxKey) {
  jsonlite::fromJSON(UrlDecaux(decaux,key), flatten = TRUE)
}

Contracts <- GetJsonDecaux("contracts")
DecauxContractName <- filter(Contracts, commercial_name == "Velib")[["name"]]
Stations <- GetJsonDecaux(sprintf("stations?contract=%s", DecauxContractName))
Stations <- mutate(Stations, status = factor(status, level=c("CLOSED","OPEN")))
Stations <- mutate(Stations, contract_name = factor(contract_name))
Stations <- mutate(Stations, date = as.POSIXct(last_update/1000, origin = "1970-01-01"))
StationsDate <- max(Stations[,'date'])

location.lat.max <- max(Stations[["position.lat"]])
location.lat.min <- min(Stations[["position.lat"]])
location.lat.width <- location.lat.max-location.lat.min
location.lng.max <- max(Stations[["position.lng"]])
location.lng.min <- min(Stations[["position.lng"]])
location.lng.width <- location.lng.max-location.lng.min
location.box <- c(location.lat.min-.05*location.lat.width,
                  location.lat.max+.05*location.lat.width,
                  location.lng.min-.05*location.lng.width,
                  location.lng.max+.05*location.lng.width)
names(location.box) <- c("bottom", "top", "left", "right")

library(ggmap)
map.Decaux.raw <- get_map(location.box, source = "google", 
                  maptype = "roadmap")
map.Decaux <- ggmap(map.Decaux.raw, extent = "device")

map.avail <- map.Decaux + geom_point(data = Stations, 
                       aes(x = position.lng, y = position.lat, 
                           col = available_bikes/bike_stands, 
                           size = bike_stands),
                       alpha = .85) + 
  scale_size_continuous(range = c(.5,3), name = "Bike stands") +
  scale_color_gradient(limits = c(0,1), name = "Bike availability") +
  ggtitle("Bike availability") +
  theme(plot.title = element_text(size = 20))
map.avail
```

# Density and contours

```{r Density_2D}
StationsRep <- Stations[rep(1:nrow(Stations), Stations$bike_stands),]

map.Decaux + 
  stat_density_2d(data = StationsRep, 
                 aes(x = position.lng, y = position.lat, 
                     alpha = ..level.. ,  fill = ..level..), 
                 contour = TRUE, geom = "polygon") +  scale_alpha_continuous(guide = "legend") + scale_fill_continuous(guide = "legend")
```

# Cartogram

```{r Cartogram}
library(GISTools)
library(Rcartogram)
library(getcartr)
data(georgia)
georgia.carto <- quick.carto(georgia,georgia$TotPop90)
georgia.carto_fortify <- left_join(fortify(georgia.carto, region = "Name"), dplyr::select(georgia.carto@data, Name, TotPop90), by = c("id" = "Name"))
ggplot(data = georgia.carto_fortify, aes(x = long, y = lat, group = group, fill = TotPop90)) + geom_polygon()

georgia_fortify <- left_join(fortify(georgia, region = "Name"), dplyr::select(georgia@data, Name, TotPop90), by = c("id" = "Name"))
ggplot(data = georgia_fortify, aes(x = long, y = lat, group = group, fill = TotPop90)) + geom_polygon()
```

# Tree

```{r Tree}
library(rpart)
library(rpart.plot)

data(airquality)
airq <- subset(airquality, !is.na(Ozone))
airct <- rpart(Ozone ~ ., data = airq, control = rpart.control(minsplit = 10))

rpart.plot(airct)
prp(airct, type = 2, extra = 1, nn = TRUE)
prp(airct, type = 2, extra = 1, nn = TRUE, fallen.leaves = TRUE)
```

# Tree Graph

```{r Tree_Graph}
library(treemap)
data(business)
treegraph(business, index=c("NACE1", "NACE2", "NACE3", "NACE4"), show.labels=FALSE)
```

# Graph

```{r Graph}
library("igraph")
mis_file = "lesmiserables.txt"
mis_graph = read.graph(mis_file, format = "gml")
mis_graph = permute.vertices(mis_graph, order(V(mis_graph)$group))

plot(mis_graph, vertex.size = 5, edge.width = E(mis_graph)$value, vertex.color = V(mis_graph)$group)

plot(mis_graph, layout = layout.fruchterman.reingold, vertex.size = 5, edge.width = E(mis_graph)$value, vertex.color = V(mis_graph)$group)

plot(mis_graph, layout = layout.circle, vertex.size = 5, edge.width = E(mis_graph)$value, vertex.color = V(mis_graph)$group)
```

```{r visNetwork, cache = FALSE}
library(visNetwork)

Nodes <- igraph::get.data.frame(mis_graph, what = "vertices")
Edges <- igraph::get.data.frame(mis_graph, what = "edges")
Edges <- dplyr::mutate(Edges, from = Nodes[from,"id"], to = Nodes[to,"id"])

visNetwork(nodes = Nodes,
           edges = Edges) %>%
  visOptions(highlightNearest = TRUE) %>%
  visPhysics(solver = "barnesHut")
```

# Arc diagram

```{r Arcdiagram}
library(arcdiagram)
edgelist = get.edgelist(mis_graph)
vlabels = get.vertex.attribute(mis_graph, "label")
vgroups = get.vertex.attribute(mis_graph, "group")
vfill = get.vertex.attribute(mis_graph, "fill")
vborders = get.vertex.attribute(mis_graph, "border")
degrees = degree(mis_graph)
values = get.edge.attribute(mis_graph, "value")
library(reshape)
x = data.frame(vgroups, degrees, vlabels, ind = 1:vcount(mis_graph))
y = arrange(x, desc(vgroups), desc(degrees))
new_ord = y$ind
arcplot(edgelist, ordering = new_ord, labels = vlabels, cex.labels = 0.8,
        show.nodes = TRUE, col.nodes = vborders, bg.nodes = vfill,
        cex.nodes = log(degrees) + 0.5, pch.nodes = 21, lwd.nodes = 2, line = -0.5,
        col.arcs = hsv(0, 0, 0.2, 0.25), lwd.arcs = 1.5 * values)
```

#  Adjacency Matrix

```{r Adjacency_Matrix}
edges <- data.frame( from = edgelist[,1], to = edgelist[,2], values = values)
edgessym <- data.frame( from = edgelist[,2], to = edgelist[,1], values = values)
edges <- unique(rbind(edges, edgessym))

ggplot(data = edges, aes(x = factor(from, labels = vlabels), 
                         y = factor(to, labels = vlabels), 
                         color = log(values+1))) +
  geom_raster() + xlab("") + ylab("") +
  theme(axis.text.x = element_text(angle = 270, hjust = 0))
```

# Flow

```{r Flow}
library(riverplot)
data(minard )
nodes <- minard$nodes
edges <- minard$edges
colnames( nodes ) <- c( "ID", "x", "y" )
colnames( edges ) <- c( "N1", "N2", "Value", "direction" )

# color the edges by troop movement direction
edges$col <- c( "#e5cbaa", "black" )[ factor( edges$direction ) ]
# color edges by their color rather than by gradient between the nodes
edges$edgecol <- "col"

# generate the riverplot object and a style
river <- makeRiver( nodes, edges )
style <- list( edgestyle= "straight", nodestyle= "invisible" )

# plot the generated object
plot( river, lty= 1, default_style= style )
# Add cities
with( minard$cities, points( Longitude, Latitude, pch= 19 ) )
with( minard$cities, text( Longitude, Latitude, Name, adj= c( 0, 0 ) ) )
```

# 3D

```{r 3D, webgl = TRUE}
library(rgl)

plot3d(mtcars[c(3,6,1)])

plot3d(small_diamonds[c(1,2,7)])
```


# Animation

```{r Animation, fig.show = 'animate'}
for (year in names(st)[7:37]) { 
plot(ggplot(filter(st_melt, variable == year), aes(long, lat, group=group, fill = value)) +
    geom_polygon() + scale_fill_continuous(limits = c(20,80)) + facet_wrap(~ variable))
}
```

# Javascript

```{r GoogleVis, results='asis'}
# Data source: http://goo.gl/vcKo6y
UKvisits <- data.frame(origin=c(
  "France", "Germany", "USA",
  "Irish Republic", "Netherlands",
  "Spain", "Italy", "Poland",
  "Belgium", "Australia", 
  "Other countries", rep("UK", 5)),
  visit=c(
    rep("UK", 11), "Scotland",
    "Wales", "Northern Ireland", 
    "England", "London"),
  weights=c(
    c(12,10,9,8,6,6,5,4,4,3,33)/100*31.8, 
    c(2.2,0.9,0.4,12.8,15.5)))
## Uncomment the next 3 lines to install the developer version of googleVis
# install.packages(c("devtools","RJSONIO", "knitr", "shiny", "httpuv"))
# library(devtools)
# install_github("mages/googleVis")
require(googleVis)
plot(
  gvisSankey(UKvisits, from="origin", 
             to="visit", weight="weight",
             options=list(
               height=250,
               sankey="{link:{color:{fill:'lightblue'}}}"
               )),
  tag = "chart"
)

```

```{r Plotly}
library(plotly)
gg <- ggplot(data = mtcars, aes(x = hp, y = mpg, size = gear, color = cyl, shape = factor(am))) + geom_point() + scale_size_continuous(range = c(4,8))
```

```{r Plotly2, cache = FALSE}
library(plotly)
ggplotly(gg)
```

```{r Bokeh, cache = FALSE}
library(rbokeh)
p <- figure() %>% ly_points(data = dplyr::mutate(mtcars, cyl = factor(cyl)), x = hp, y = mpg, color = cyl, glyph = factor(am), hover = mtcars)
p
```

```{r Leaflet, cache = FALSE}
library(leaflet)

ColorPal <- colorNumeric(scales::seq_gradient_pal(low = "#132B43", high = "#56B1F7", space = "Lab"), domain = c(0,1))
m <- leaflet(data = Stations) %>%
  addTiles() %>%
  addCircles(~ position.lng, ~ position.lat, popup = ~ sprintf("<b> Available bikes: %s</b>",as.character(available_bikes)),
                   radius = ~ sqrt(bike_stands),
            color = ~ ColorPal( available_bikes / (available_bikes + available_bike_stands)),
            stroke = TRUE, fillOpacity = 0.75)
m
```

# Big Data

```{r BigData}
airline<-read.csv("airlineJan.csv")
head(airline)
nrow(airline)
dim(airline)

library(bigvis)
library(ggplot2)
detach("package:dplyr")
library(plyr)
library(dplyr)
library(grid)
library(reshape2)
library(scales)
library(memoise)
library(hexbin)


delay <- airline$ArrDelay
dist <- airline$Distance
time <- airline$AirTime

speed <- dist / time * 60

ggplot(data = airline, aes(x = dist, y = time)) + geom_point()
hexbinplot(time~dist)
```

